---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Multinomial classification of turtles

<!-- badges: start -->
<!-- badges: end -->

This GitHub repository contains the code used in analysis of all datasets used in __Cold-adapted Cretaceous polar turtle resolves the trans-Antarctic radiations of Testudinata__

by Benjamin P. Kear, Márton Rabi, Martin Kundrát, Michael S. Y. Lee, Daniel Snitting,
Peter Trussler, Mohamad Bazzi, Barbara E. Wagstaff, Dorris E. Seegets-Villiers, Thomas H. Rich, Patricia Vickers-Rich & Lesley Kool

Code written and maintained by Mohamad Bazzi
Contact: bazzi@ug.kth.se

## Access .Rdata

Larger assets can be accessed from within a report using [`piggyback`](https://github.com/ropensci/piggyback)!

```{r}
# Install and load R package
install.packages(piggyback)
require(piggyback)

# Create temporar directory and load .Rdata into R environment,
pb_download(file = "default.RData",dest = tempdir(),repo = "mohabazzi/Turtle-Project-2023",tag = "v1.0.0")
load(file = file.path(tempdir(),"default.RData"))
```
**Data files included:**

Measurement File.csv - Extant and fossil testudinatan forelimb proportion dataset of Joyce & Gauthier.

